TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-09 src/paper.tex
update

The discretization of audio has gained significant attention with the advancement of language models (LMs). It has been explored in
tasks such as speech enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}.
This method converts audio into discrete tokens, and leverages LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. SSL models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performances across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech. As shown in \cite{dasb}, SSL models outperform neural audio codecs \cite{dac} in tasks like speech enhancement and separation. Consequently, this paper primarily focuses on exploring the discretization of SSL models.

The discretization approach for Target Speech Extraction (TSE) has received limited attention in the literature. One of the earliest works to introduce discrete tokens in the context of TSE is presented by \cite{gen_tse}, which investigates the use of self-supervised learning (SSL) models such as HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, this study does not explore the WavLM SSL model, which has shown superior performance in speech separation tasks \cite{wavlm}. Additionally, it primarily utilizes the output from a single layer of the SSL model for discretization. Furthermore, the evaluation in \cite{gen_tse} focuses solely on speech quality, neglecting key metrics such as speech intelligibility and speaker similarity, both of which are critical for the TSE task. To address these limitations, we propose TSELM, a novel approach that leverages multiple hidden layers of WavLM for discretization and incorporates evaluation metrics for speech intelligibility and speaker similarity.

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech. It serves as the upper bond for our model.  Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross attention  
and LM without discretization. 
The concatenation strategy is stilled applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but without concatenation strategy.

@@

  \caption{The performance of different systems on Libri2Mix testset. For TSELM-L-Hybrid  

@@

  Mixture                 & -                                             & -                                                                                           & 3.39        & 3.14       & 2.68       & 79.3            & -        &           & 3.39        & 3.14       & 2.68       & 79.3            & -                  \\
  Target-Discrete         & G                                             & -                                                                                          & 3.57        & 4.10       & 3.32       & 11.3            & 0.653     &          & 3.57        & 4.10       & 3.32       & 11.3            & 0.653               \\ \hline
  Spex+                   & D                                             & -                                                                                  & 3.36        & 3.76       & 2.98       & 19.3            & 0.923     &          & 3.36        & 3.76       & 2.98       & 19.3            & 0.923             \\ \hline
  Continuous-WavLM-L6     & G                                             & C                                                                                  & 3.56        & 4.06       & 3.27       & 14.7            & 0.877     &          & 3.56        & 4.06       & 3.27       & 14.7            & 0.877             \\
  TSELM-L-Hybrid          & G                                             & H                                                                                     & 3.50        & 4.06       & 3.22       & 19.8            & 0.924\_d  &          & 3.50        & 4.06       & 3.22       & 19.8            & 0.924\_d             \\
  TSELM-S-NoCat       & G                                             & D                                                                                   & 3.47        & 4.03       & 3.19       & 69.6            & 0.868\_d    &        & 3.47        & 4.03       & 3.19       & 69.6            & 0.868\_d         \\ \hline
  TSELM-S                 & G                                             & D                                                                                    & 3.50        & 4.07       & 3.24       & 28.1            & 0.892\_d      &      & 3.50        & 4.07       & 3.24       & 28.1            & 0.892\_d            \\
  TSELM-M                 & G                                             & D                                                                                   & 3.49        & 4.05       & 3.22       & 28.4            & 0.901\_d    &        & 3.49        & 4.05       & 3.22       & 28.4            & 0.901\_d             \\
  TSELM-L                 & G                                             & D                                                                                  & 3.49        & 4.05       & 3.22       & 27.1            & 0.905\_d    &        & 3.49        & 4.05       & 3.22       & 27.1            & 0.905\_d        \\

@@

\section{Results and Discussions}
Table \ref{main_exp} presents the performance of different systems evaluated on the Libri2Mix test.
The model size for Target-Discrete refers to the total size of the encoder and 
decoder, which are WavLM Large and scalable HiFi-GAN, respectively. DNSMOS is computed over the 
testset since this metric is reference-free. dWER is calculated with respect to the clean speech. 
For continuous methods, speaker similarity is directly computed against the clean speech. 
However, since discretization inherently results in some loss of speaker information—Target-Discrete shows a speaker similarity score of 0.653 compared to continuous methods—we assess the speaker similarity of the output from the discrete methods against the target audio produced by discretizing the clean speech. 
This target audio output from Target-Discrete serves as the upper bond for our model performance.
The observed speaker information loss is likely due to the tokenization process, which inherently reduces speaker fidelity. Future work should aim to enhance tokenization methods for SSL models to mitigate this loss. Since the primary goal of this research is to explore target speaker extraction using discretized information rather than to develop improved tokenization methods, we argue that comparing the outputs of our model with discretized speech is a reasonable approach, as it represents an upper bound on performance.

@@

We observe a significant increase in dWER when the mixture audio is not concatenated with the reference, as seen in the TSELM-S-NoCat results in Table \ref{main_exp}. For WavLM to effectively perform target speaker separation, the input audio must follow specific conditions: the mixture should be less than 50\% of the total length, and the first utterance should be the target speaker. Under these conditions, WavLM outputs a slightly denoised embedding that emphasizes the target speaker. When the entire input is a mixture, however, we found that WavLM sometimes extracts the wrong speaker.
Our current concatenation strategy is inspired by SELM's \cite{selm} success in speech denoising and aims to transform the target speaker extraction task into a more difficult speech enhancement problem leveraging WavLM's denoising capability.

>>>>

The discretization of audio has gained significant attention with the advancement of language models (LMs). It has been explored in
tasks such as speech enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}.
This method converts audio into discrete tokens, and leverages LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. SSL models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performances across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech.

Discretization in Target Speech Extraction (TSE) has seen limited study. \cite{gen_tse} is among the first to use discrete tokens in TSE, leveraging SSL models like HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, it overlooks the WavLM model, which excels in speech separation, and focuses on single-layer outputs for discretization. Moreover, its evaluation is limited to speech quality, neglecting intelligibility and speaker similarity, both crucial for TSE. We address these gaps by proposing TSELM, which uses multiple WavLM layers for discretization and adds metrics for intelligibility and speaker similarity.

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
We also compare our model with UniCATS(vq-wav2vec) from \cite{gen_tse}, which uses discrete tokens from wav2vec to conduct TSE. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech. 
It serves as the upper bond for our model.  Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross attention  
and LM without discretization. 
The concatenation strategy is stilled applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but without concatenation strategy.

@@

  \caption{Results on Libri2Mix testset and wsj0\_2mix\_extra testset. For TSELM-L-Hybrid  

@@

  Mixture                 & -                                             & -                                                                                           & 3.39        & 3.14       & 2.68       & 79.3            & -        &           & 3.42        & 3.28       & 2.81       & 63.6            & -                  \\
  Target-Discrete         & G                                             & -                                                                                          & 3.57        & 4.10       & 3.32       & 11.3            & 0.653     &          & 3.56        & 4.09       & 3.30       & 10.1            & 0.657               \\ \hline
  Spex+                   & D                                             & -                                                                                  & 3.36        & 3.76       & 2.98       & 19.3            & 0.923     &          & 3.49        & 4.00       & 3.21       & 15.0            & 0.943             \\
  SkiM-UniCATS                   & G                                             & D                                                                                  & -        & -      & -       & -            & -     &          & 3.62        & 4.10       & 3.37       & -            & -             \\ \hline
  Continuous-WavLM-L6     & G                                             & C                                                                                  & 3.56        & 4.06       & 3.27       & 14.7            & 0.877     &          & 3.61        & 4.08       & 3.35       & 8.0            & 0.892             \\
  TSELM-L-Hybrid          & G                                             & H                                                                                     & 3.50        & 4.06       & 3.22       & 19.8            & 0.924\_d  &          & 3.57        & 4.10       & 3.31       & 12.6            & 0.915\_d             \\
  TSELM-S-NoCat       & G                                             & D                                                                                   & 3.47        & 4.03       & 3.19       & 69.6            & 0.868\_d    &        & 3.55        & 4.08       & 3.28       & 64.5            & 0.888\_d         \\ \hline
  TSELM-S                 & G                                             & D                                                                                    & 3.50        & 4.07       & 3.24       & 28.1            & 0.892\_d      &      & 3.57        & 4.10       & 3.32       & 19.4            & 0.915\_d            \\
  TSELM-M                 & G                                             & D                                                                                   & 3.49        & 4.05       & 3.22       & 28.4            & 0.901\_d    &        & 3.57        & 4.10       & 3.32       & 18.8            & 0.921\_d             \\
  TSELM-L                 & G                                             & D                                                                                  & 3.49        & 4.05       & 3.22       & 27.1            & 0.905\_d    &        & 3.57        & 4.10       & 3.31       & 17.8            & 0.924\_d        \\

@@

\section{Results and Discussions}
Table \ref{main_exp} presents the performance of different systems evaluated on the Libri2Mix test and wsj0\_2mix\_extra testset. 
DNSMOS is computed over the 
output since this metric is reference-free. dWER is calculated with respect to the clean speech. 
For continuous methods, speaker similarity is directly computed against the clean speech. 
However, since discretization inherently results in some loss of speaker information—Target-Discrete shows a speaker similarity score of 0.653 compared to continuous methods—we assess the speaker similarity of the output from the discrete methods against the target audio produced by discretizing the clean speech. 
This target audio output from Target-Discrete serves as the upper bond for our model performance.
The observed speaker information loss is likely due to the tokenization process, which inherently reduces speaker fidelity. Future work should aim to enhance tokenization methods for SSL models to mitigate this loss. Since the primary goal of this research is to explore target speaker extraction using discretized information rather than to develop improved tokenization methods, we argue that comparing the outputs of our model with discretized speech is a reasonable approach, as it represents an upper bound on performance.

@@

We observe a significant increase in dWER when the mixture audio is not concatenated with the reference, as seen in the TSELM-S-NoCat results in Table \ref{main_exp}. For WavLM to effectively perform target speaker separation, the input audio must follow specific conditions: the mixture should be less than 50\% of the total length, and the first utterance should be the target speaker. Under these conditions, WavLM outputs a slightly denoised embedding that emphasizes the target speaker. When the entire input is a mixture, however, we found that WavLM sometimes extracts the wrong speaker.
Our current concatenation strategy is inspired by SELM's \cite{selm} success in speech denoising and aims to transform the target speaker extraction task into a more difficult speech enhancement problem leveraging WavLM's denoising capability. However, this is a suboptimal solution and we think future work should focus on developing a speaker-aware tokenization method.
